Overview and Summary of the First Workshop on Sustainable Software for Science: Practice and Experiences (WSSSPE)

### 2014-02-20 WSSSPE1_report/WSSSPE1_summary_report.tex
small updates and smoothing in sections 1 and 9

Because progress in scientific research is dependent on the quality and
accessibility of software at all levels, it is now critical to address many
challenges related to the development, deployment, and maintenance of reusable
software.
%\note{Ketan: What are the levels in scientific process where software
%is required. It might be interesting to outline these stages and types of
%software required in each stage.}
In addition, it is essential that scientists,
researchers, and students are able to learn and adopt software-related skills
and methodologies. Established researchers are already acquiring some of these
skills, and in particular a specialized class of software developers is
emerging in academic environments who are an integral and embedded part of
successful research teams. The WSSSPE workshop was intended to provide a forum
for discussion of the challenges, including both positions and experiences. The
short papers and discussion have been archived to provide a basis for continued
discussion, and the workshop has fed into the collaborative writing of this
journal publication. An estimate of 90 to 150 paricipants were present 
throughout the day. Additional papers based on extended versions of workshop
submissions are expected.  The level of interest in the workshop has led the
organizers, working with some of the submitters and attendees, to plan two
additional workshops (at the 2014 SciPy and SC conferences) and to turn the
workshop website into a community website that can be used as a focus for
further discussion and progress.  A minisymposium on ``Reliable Computational
Science'' that aims to further explore some of the key issues raised in WSSSPE
is co-organized by a participant in the 2014 SIAM Annual Meeting.

@@

For the purpose of studying software case-studies we classify the software
projects discussed in the workshop into three broad categories. First, the
`utility software', software tools that enable and/or facilitate the
development of other tools and techniques to carry out scientific work. Second,
the `scientific software', software that was developed with an aim to solve a
class of scientific problems. Third, the `infrastructure software', software
developed to efficiently utilize new research infrastructure.

Although we anticipate overlap in categories and point it out in our
discussion. The classification is based on two criterions: First, the character
of a software project that dominates and best fits and the theme the paper
discussing the project was placed in the workshop discussion.

We discuss the case studies with an aim of understanding the practicalities of
the many points discussed in the workshop and hope to extract the best
practices and associate them with the projects. We discuss how the cases are
applicable to and has impacted user communities over their lifetime isolating
similarities and differences between projects. What are the takeaways for a new
project and the lessons learned?  How can the course of current and
future development projects be altered to implement lessons learned.

\subsection{Utility Software}
The General Purpose Software applicable to a generic task and/or enabling other
software to run by providing suitable environment. For example, collaborative
development, Workflow and scripting frameworks, visualization frameworks.
\subsection{Scientific Software}
Special Purpose Software that was developed specifically for a target use-case
scenario or keeping some fixed/frozen requirements in mind.  
Domain science
oriented software developed to solve domain science problems, eg MINRES-QLP.
\subsection{Infrastructure Software}
Infrastructure oriented software developed to better utilize new
infrastructure, eg. MVAPICH.
%
%\begin{itemize}
%  \item mvapich, Cactus, DUNE
%  \item gateways, github, VisIT, visTrails, PAAS
%  \item SimGrid, Swift, Galaxy
%  \item FastMath, R/qtl, MINRES-QLP
%  \item OWL/RDF, AMR/multiphysics, Globus-Genomics
%\end{itemize}
%

\note{Dan: do we need this section?  I think there's some overlap
  between this and the discussion in the Developing and Support
  Software section.  Or perhaps this should go in the cross-cutting
  issues section?  Let's write it down here - then we can decide if it goes elsewhere}
\note{Ketan: yes, may be this section will split and merge into other sections}

>>>>

Because progress in scientific research is dependent on the quality and
accessibility of software at all levels, it is now critical to address many
challenges related to the development, deployment, and maintenance of reusable
software.
%\note{Ketan: What are the levels in scientific process where software
%is required. It might be interesting to outline these stages and types of
%software required in each stage.}
In addition, it is essential that scientists,
researchers, and students are able to learn and adopt software-related skills
and methodologies. Established researchers are already acquiring some of these
skills, and in particular a specialized class of software developers is
emerging in academic environments who are an integral and embedded part of
successful research teams. The WSSSPE workshop was intended to provide a forum
for discussion of the challenges, including both positions and experiences. The
short papers and discussion have been archived to provide a basis for continued
discussion, and the workshop has fed into the collaborative writing of this
journal publication. An estimate of 90 to 150 participants were present 
throughout the day. Additional papers based on extended versions of workshop
submissions are expected.  The level of interest in the workshop has led the
organizers, working with some of the submitters and attendees, to plan two
additional workshops (at the 2014 SciPy and SC14 conferences) and to turn the
workshop website into a community website that can be used as a focus for
further discussion and progress.  Additionally, a minisymposium at 2014 SIAM
Annual Meeting on ``Reliable Computational
Science'' that aims to further explore some of the key issues raised in WSSSPE
is co-organized by a WSSSPE1 participant.

@@

For the purpose of studying software case studies, we classify the software
projects discussed in the workshop in three broad categories. First, 
`utility software', software tools that enable and/or facilitate the
development of other tools and techniques to carry out scientific work. Second,
`scientific software', software that was developed with an aim to solve a
class of scientific problems. Third, `infrastructure software', software
developed to efficiently utilize new research infrastructure.

These categories have some overlap, and in some \note{most?} cases, the
case studies don't cleanly fit into a single category.
Our classification is based on two criteria: first, the character
of a software project that dominates and best fits, and, second,
the theme of the workshop discussion in which the paper discussing
the project was placed.

%and point it out in our
%discussion.

Our discussion of the case studies is aimed at understanding the practicalities of
the many points discussed in the workshop, and we hope to extract the best
practices and associate them with the projects. We discuss how the cases are
applicable to and have impacted user communities over their lifetime, isolating
similarities and differences between projects. We try to define the takeaways for a new
project and the lessons learned. We try to understand how the course of current and
future development projects might be altered to implement lessons learned.

\subsection{Utility Software}
Utility software potentially has the broadest community impact of the three
categories. This is
especially relevant in the modern era of social computing. In the context of
this paper, a utility software project could be defined as general purpose software
applicable to one or many generic tasks and/or enabling other software to run
by providing a suitable environment. Examples are collaborative development
framework such as \emph{GitHub} and \emph{Bitbucket}, Workflow and scripting
frameworks such as \emph{Galaxy}, \emph{Swift} and \emph{VisTrails},
and visualization frameworks such as \emph{visIT}. Often projects start with
specific, small set of requirements and users in mind but expand to provide
more generic functionality and appeal to a wider user community.

\subsection{Scientific Software}
Special purpose software that was developed specifically for a target use-case
scenario or keeping some fixed/frozen requirements in mind.  Domain science
oriented software developed to solve domain science problems, e.g., MINRES-QLP.

\subsection{Infrastructure Software}
Infrastructure oriented software developed to better utilize new
infrastructure, e.g., MVAPICH.
%
%\begin{itemize}
%  \item mvapich, Cactus, DUNE
%  \item gateways, github, VisIT, visTrails, PAAS
%  \item SimGrid, Swift, Galaxy
%  \item FastMath, R/qtl, MINRES-QLP
%  \item OWL/RDF, AMR/multiphysics, Globus-Genomics
%\end{itemize}
%

\note{Dan: I think there's some overlap
  between this and the discussion in the Developing and Support
  Software section.  Perhaps this should go in the cross-cutting
  issues section?  Let's write it down here - then we can decide if it goes elsewhere}
\note{Ketan: yes, may be this section will split and merge into other sections}
